---
title: "Converting Match Scores to Likelihoods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Converting Match Scores to Likelihoods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Purpose

TaxaLikely converts raw match scores (DNA percent identity, image similarity,
acoustic confidence) into calibrated statistical likelihoods for taxonomic
assignment. It sits between TaxaMatch (which produces match scores) and
TaxaAssign (which combines likelihoods with priors to compute posteriors).

**The core idea:** A 98% match to Species A means different things depending on
the database. If Species A typically matches at 99.5%, a 98% match is
suspicious. If it typically matches at 96%, a 98% match is excellent. TaxaLikely
learns these species-specific patterns from a reference database.

## Quick Start

```{r quickstart}
library(TaxaLikely)

# 1. Build reference alignment matrix
ref_matrix <- build_sequence_matrix(reference_df)

# 2. Train the likelihood model
model <- train_likelihood_model(ref_matrix)

# 3. Apply to match data
# ratio_threshold = 0: gate H1 rows by the score-only outlier filter (alpha = 0.001)
# rather than by comparison to H2/H3 likelihoods.
result <- evaluate_likelihoods(match_df, model_params = model, ratio_threshold = 0)
likelihoods <- result$likelihoods
```

## Detailed Walkthrough

### Step 1: Acquire reference sequences

TaxaLikely provides two ways to build a `reference_df`:

```{r fetch-refs}
# Option A: Fetch from NCBI by taxon + barcode marker
reference_df <- fetch_reference_sequences(
 taxa         = c("Fundulidae", "Atherinopsidae", "Gobiidae"),
 barcode_term = "12S",
 max_per_species = 5,
 cache_dir       = "ncbi_cache"
)

# Option B: Read from a local FASTA + taxonomy table
reference_df <- read_reference_fasta(
 fasta_path  = "my_references.fasta",
 taxonomy    = taxonomy_table,
 rank_system = c("family", "genus", "species")
)
```

### Step 2: Build the reference matrix

Align all reference sequences against each other to learn the expected score
distributions.

```{r build-matrix}
ref_matrix <- build_sequence_matrix(reference_df)
# Returns a data frame with columns: query, subject, p_match, family, genus, species
```

### Step 3: Check for mislabeled references

Before training, identify potential errors in the reference database.

```{r flag-errors}
errors <- flag_reference_errors(ref_matrix)
# Inspect flagged sequences
print(errors)

# Remove flagged sequences from the match object before evaluation
match_df_clean <- match_df[!match_df$accession %in% errors$accession, ]
```

### Step 4: Train the likelihood model

```{r train}
model <- train_likelihood_model(ref_matrix)
print(model)  # Summary: n_species, AIC, global parameters
```

The model learns three hypothesis distributions:

- **H1 (correct species):** Expected score when query truly belongs to this
  species. Per-species parameters with Empirical Bayes shrinkage.
- **H2 (unreferenced species):** Expected score when the true species is absent
  from the reference — a congener exists but the actual species does not.
- **H3 (unreferenced genus):** Expected score when even the genus is absent.

### Step 5: Evaluate likelihoods

```{r evaluate}
result <- evaluate_likelihoods(
  match_df        = match_df,
  model_params    = model,
  rank_system     = c("family", "genus", "species"),
  n_sims          = 1000,
  ratio_threshold = 0    # recommended: rely on score-only outlier filter (alpha = 0.001)
)

likelihoods <- result$likelihoods
unresolved  <- result$unresolved
# unresolved: observations with no usable likelihoods (retry with coarser rank_system)
```

### Step 6: Filter and constrain

```{r filter}
# Keep only the finest-rank candidates per observation
top <- filter_top_hypotheses(likelihoods, rank_system = c("family", "genus", "species"))

# Audit reference completeness
coverage <- audit_barcode_coverage(
 match_df     = match_df,
 barcode_term = "12S",
 target_rank  = "genus"
)

# Suppress unreferenced species for fully-sampled genera
constrained <- apply_coverage_constraints(top, coverage$census)
```

The output is ready for TaxaAssign: `expand_unreferenced_hypotheses()` →
`join_priors()` → `compute_posterior()`.

## Key Parameters

| Parameter | Default | When to change |
|---|---|---|
| `n_sims` | 1000 | Set to 0 for point estimates only (faster) |
| `ratio_threshold` | 0.01 | **Set to 0** for production workflows. The default compares H1 density to H2/H3 density (a cross-rank comparison) and can suppress legitimate H1 candidates in species-rich families. With `ratio_threshold = 0`, H1 filtering relies solely on the score-only outlier check (`alpha`) |
| `alpha` | 0.001 | H1 score-outlier threshold (~3.3 sigma, df=1 chi-sq, score only). Increase to admit weaker H1 matches; decrease to tighten. Gap is excluded from this test — small gaps (confusable congeners) are handled by the bivariate density, not the outlier filter |
| `anchor_perfect` | TRUE | Set FALSE if perfect matches are impossible |
| `barcode_term` | — | Must match your marker: "12S", "COI", "ITS2", etc. |
| `mislabel_threshold` | — | Lower to flag more borderline references |

## Common Pitfalls

1. **"No usable likelihoods" for some observations:** The `$unresolved` component
   contains these. Usually caused by all candidates matching only at ranks
   coarser than `rank_system`. Re-run with a broader rank system.

2. **DECIPHER not installed:** `build_sequence_matrix()` requires the
   Bioconductor package DECIPHER. Install with
   `BiocManager::install("DECIPHER")`.

3. **Score scale confusion:** Match scores can be 0–100 (percent) or 0–1
   (proportion). TaxaLikely auto-detects and normalises, but check if results
   look wrong.

4. **Singleton species:** Species with only one reference sequence get
   conservative (shrunk-to-global) parameters. This is by design — limited
   data means more uncertainty.
