---
title: "Building Occurrence-Based Priors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building Occurrence-Based Priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, purl = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  purl = FALSE
)
```

<!--
  EVERY chunk below carries `eval = FALSE, purl = FALSE` in its own header, and
  that repetition is load-bearing -- do not "DRY it up" into the setup chunk.
  R CMD check's "running R code from vignettes" step TANGLES the vignette with
  knitr::purl(), and purl() does NOT execute the setup chunk, so an
  opts_chunk$set(purl = FALSE) written there has no effect at tangle time: every
  chunk still lands in the tangled .R file and is sourced. This vignette is
  documentation-only (its code makes live GBIF/NCBI calls and references
  objects its prose never defines, e.g. `occurrences`), so a tangled chunk is a
  guaranteed check ERROR. This exact bug hit this vignette once already
  (fixed 2026-09-03, on a since-removed quickstart chunk) -- see
  `TaxaExpect/CLAUDE.md`'s 2026-09-03 session note for the full record.
-->

## What TaxaExpect Computes, and Why

A taxonomic assignment pipeline usually has to choose among several
candidate species that all match the evidence about equally well. Two
mockingbird-sized shorebirds, two look-alike gobies, a fish and its
introduced-but-absent-here relative: the sequence, image, or acoustic
match alone often can't tell them apart. What can help is a second,
independent piece of information: how likely is each candidate to
actually occur at this location?

That's what TaxaExpect computes: a **prior**, one number per candidate
species, expressing how plausible it is that a random legitimate
detection at a site belongs to that species, before any match evidence
is considered. Combined with a likelihood (from TaxaLikely) via Bayes'
theorem, a well-calibrated prior can break a tie between two
equally-scoring candidates in favor of the one that's actually known to
live there, and can down-weight a candidate that scores well but has
never been recorded anywhere near the site.

TaxaExpect builds these priors from occurrence records, typically GBIF
observations fetched by TaxaFetch and habitat-labelled by TaxaHabitat
(though any similarly-shaped data works). It estimates a species' prior as
its distance-weighted share of nearby occurrence records: records close
to the site count fully, records farther away count less, and the decay
is smooth rather than an abrupt cutoff. Even a site with very few nearby
records gets a usable prior, because it borrows some information from
the surrounding region rather than falling back to nothing.

This vignette walks through that estimation process end to end: choosing
how quickly influence should decay with distance, estimating a prior at
a real site, handling species that were never recorded at all, and
producing a final table ready to hand to TaxaAssign.

## Quick Start

The shape of a full run, before the walkthrough below explains each
step:

```{r quickstart, eval = FALSE, purl = FALSE}
library(TaxaExpect)

# 1. Calibrate the kernel bandwidth (and regional back-off m) by
#    leave-one-block-out composition prediction -- never hand-set.
calib <- calibrate_kernel_bandwidth(
  occurrence_data = occurrences,   # habitat-labelled occurrence records
  site_habitat    = "Marine",
  lambda_grid     = c(10, 25, 50, 100, 200)   # km
)
lambda_km <- calib$best$lambda_km

# 2. Estimate kernel priors at your site (resident_observed rows)
kernel_fit <- estimate_kernel_priors(
  occurrence_data = occurrences,
  site_lat        = 34.45,
  site_lon        = -120.47,
  site_habitat    = "Marine",
  lambda_km       = lambda_km
)

# 3. Add resident_undetected rows (singleton mirrors + Good-Turing floor)
undetected <- generate_undetected_diversity(kernel_fit)

# 4. Assemble the prior table for TaxaAssign
priors <- dplyr::bind_rows(kernel_fit$priors, undetected)

# 5. Explore the prior field
plot_theta_surface(kernel_fit, occurrence_data = occurrences,
                    taxon = "Girella nigricans")
```

The rest of this vignette explains each of these steps, plus two things
the Quick Start skips: what to do about species your data never
recorded at all, and a correctness issue (mixing incompatible detection
methods) that a first-time user needs to know about before running this
on real data.

## Step 1: Start with habitat-labelled occurrence data

`estimate_kernel_priors()` expects one row per occurrence record, with
at least a taxon name, a latitude/longitude, and a habitat category.
Fetching and labelling that data is the job of two upstream packages,
not TaxaExpect itself:

```{r fetch-habitat, eval = FALSE, purl = FALSE}
library(TaxaFetch)
library(TaxaHabitat)

keys <- get_keys_from_context(data.frame(taxon_name = candidate_taxa))
bbox <- make_bbox_wkt(lat = 34.45, lon = -120.47, radius_deg = 2)

occurrences_raw <- get_gbif_occurrences(keys = keys$usageKey, geometry = bbox)
occurrences_clean <- filter_gbif_quality(occurrences_raw)

occurrences <- assign_habitat_biological(
  occurrence_data = occurrences_clean,
  habitats_df     = habitat_lookup
)
```

The rest of this vignette assumes `occurrences` is a data frame shaped
like that: one row per record, columns for taxon name, latitude,
longitude, and a habitat category (`main_habitat` by default).

## Step 2: Choose the bandwidth (`calibrate_kernel_bandwidth()`)

The kernel estimator has one tuning value that matters most: how quickly
a record's influence decays with distance from the site (`lambda_km`).
Set it too small and the estimate rests on a handful of very close
records, ignoring real regional structure; set it too large and every
site in the study area gets nearly the same, over-smoothed answer.
There's no universal right value: it depends on how occurrence records
are distributed for your particular dataset, so TaxaExpect chooses it
empirically rather than asking you to guess.

`calibrate_kernel_bandwidth()` does this by leave-one-block-out
cross-validation: it partitions the occurrence records into spatial
blocks, predicts each held-out block's species composition from every
record outside it, and scores each candidate bandwidth by how well that
prediction matches the block's real records. It reports the bandwidth
that predicts best, alongside two references for context: what you'd
get by ignoring space entirely (`regional`), and what you'd get from the
older single-nearest-cell approach (`nearest_block`).

```{r calibrate, eval = FALSE, purl = FALSE}
calib <- calibrate_kernel_bandwidth(
  occurrence_data = occurrences,
  site_habitat    = "Marine",
  lambda_grid     = c(10, 25, 50, 100, 200)   # km, candidate bandwidths
)

calib$best             # the winning row: lambda_km, m, its score
calib$results           # every candidate scored, plus the two references
lambda_km <- calib$best$lambda_km
```

On a real Great Lakes dataset, this comparison showed why the
calibration step matters: the single-nearest-cell approach scored
*worse* than ignoring space altogether, while the distance kernel beat
both a plain regional average and the nearest-cell approach across a
wide range of bandwidths. That's an empirical result, not a stylistic
preference, and it's why TaxaExpect estimates priors this way.

If `calib$best$lambda_km` comes back at the largest value in
`lambda_grid`, `calibrate_kernel_bandwidth()` will warn you: the true
optimum may lie beyond what you tried, and it's worth widening the grid
before trusting the result. See `?calibrate_kernel_bandwidth` for
`covariate_col`/`lambda_covariate_grid` if you also want to calibrate a
covariate bandwidth at the same time, useful when a variable like depth
helps separate species that occupy the same horizontal footprint but
different depths.

## Step 3: Estimate priors at your site (`estimate_kernel_priors()`)

With a calibrated bandwidth in hand, `estimate_kernel_priors()` computes
every species' prior at one site: its kernel-weighted share of nearby
occurrence records, shrunk toward the surrounding region's own
composition so that lightly-sampled sites still get an informative
answer rather than pure noise.

```{r estimate, eval = FALSE, purl = FALSE}
kernel_fit <- estimate_kernel_priors(
  occurrence_data = occurrences,
  site_lat        = 34.45,
  site_lon        = -120.47,
  site_habitat    = "Marine",
  lambda_km       = lambda_km,
  m               = calib$best$m     # regional back-off weight, from calibration
)

kernel_fit
#> taxaexpect_kernel_priors: 214 taxa at Site_34.4500_-120.4700 (Marine)
#>   n_eff = 843 effective records (of 128301 in stratum), lambda = 25 km, m = 1
#>   singletons: 31, Good-Turing missing mass: 0.0164
#>   budget: f1 = 31, f2 = 12, chao_missing = 40, theta_present = 0.000529

head(kernel_fit$priors)
```

`kernel_fit$priors` is a table with one row per species recorded in the
focal habitat stratum, giving `alpha`/`beta` (the Beta-distribution prior
parameters TaxaAssign consumes directly), `theta_mean` (the prior
probability itself), and `effective_records` (how many effectively
independent nearby records support that estimate). A species backed by
many close records gets a tightly concentrated prior; a species with
only a few, distant, weakly-overlapping records gets a much more diffuse
one. The spread of the prior honestly reflects how much evidence went
into it.

Records are restricted to the requested habitat (`site_habitat`) before
any distance weighting happens, so a species recorded only in a
different habitat contributes nothing to this site's estimate.

## Step 4: Dark diversity for species you never detected

Every occurrence dataset is incomplete: some species that are genuinely
present at a site will have no record there at all. `estimate_kernel_priors()`
only produces rows for species that *were* recorded, so the next step
fills in a prior for everything else.

### The baseline floor

`generate_undetected_diversity()` builds two kinds of proxy rows from the
same kernel fit:

```{r undetected, eval = FALSE, purl = FALSE}
undetected <- generate_undetected_diversity(kernel_fit)
```

- **Singleton mirrors**: one anonymous proxy per species recorded
  exactly once nearby. A neighborhood with many such barely-detected
  species is telling you it likely holds more undetected diversity than
  one where every recorded species has many records.
- **A global floor**: a single low prior (`Good-Turing` missing mass
  divided across an estimated number of unrecorded species) that always
  exists, so any candidate with zero local evidence at all still gets
  *some* prior rather than an implicit zero.

These rows have no taxon name of their own: they're anonymous
placeholders representing "some undetected species," not a specific one.

### Naming a specific unobserved candidate

Sometimes you have a real, named candidate that wasn't recorded locally
but has other evidence in its favor: an occurrence just outside the
search area, a documented invasion front, a verified iNaturalist
observation nearby. For those, `apply_undetected_evidence()` prices a
*named* row instead of relying on the anonymous floor:

```{r evidence, eval = FALSE, purl = FALSE}
priors_partial <- dplyr::bind_rows(kernel_fit$priors, undetected)

evidence <- generate_presence_curve_evidence(
  taxon_names = c("Girella simplicidens"),
  w_scale     = 0.05,          # calibrated presence probability at distance 0
  distance_km = c(38)          # this species' distance to its nearest record
)

priors_with_evidence <- apply_undetected_evidence(
  taxaexpect_priors = priors_partial,
  model_obj         = kernel_fit,
  evidence          = evidence,
  grid_id           = kernel_fit$params$site_id,
  main_habitat      = "Marine",
  pricing           = "curve"
)
```

Under `pricing = "curve"`, a named claimant is priced as `theta = w *
theta_present`: with probability `w` (from the distance curve, or from
`generate_user_specified_evidence()` for a caller-asserted probability)
the species is really there, contributing `theta_present` (the typical
share of the neighborhood's own barely-detected species); with
probability `1 - w` it's absent, contributing nothing. Because
`theta_present` is pinned to the same scale as a genuinely observed
singleton, external evidence, however strong, can never make an
unrecorded species look more plausible than one your data actually
detected once.

This step is optional: skip it if you have no named unobserved
candidates worth pricing individually, and let the global floor from
`generate_undetected_diversity()` cover the rest.

## Step 5: Don't mix incompatible detection methods

Before assembling a final table, there's a correctness issue worth
understanding, not just a tuning knob: **TaxaExpect assumes every
species in a single model was detected by a comparable process.**

The estimator treats the total number of records at a site as a shared
measure of sampling effort for every species in it. That's a reasonable
assumption when every species really was sampled the same way -- e.g.
every fish species from the same eDNA marker on the same filter, or
every bird detected during one standardized point count. It breaks down
when you combine species sampled by genuinely different methods in one
model -- a phytoplankton cell count and a bird point-count sighting are
not equivalent detection events, and pooling them implies that
plankton-sampling effort says something about expected bird abundance,
which it doesn't.

**This grouping is never inferred automatically, and there is no
automated shortcut for it.** If you don't tell `estimate_kernel_priors()`
which taxa share a detection process, it has no way to know on its own
-- it silently pools everything into one shared effort denominator,
with no warning and no error. Getting this right is your responsibility
as the analyst, not something the function (or an LLM asked to guess)
can do for you: whether two taxa were detected by a comparable process
is a fact about how the data were collected, not something visible in
the taxonomy or the occurrence records themselves. An automatic,
taxonomy-based classifier for this was tried and retired after testing
against a real, expert-classified dataset: it answered a different
question than `sampling_group` is meant to answer (whether a group has
enough records at each site, not whether it shares a detection
process), splitting one real detection process into dozens of separate
automatic groups, and in one case merging two taxa the expert had kept
deliberately separate (a genuine marine survey target and likely
contamination) purely because neither had enough records on its own.

You have two options:

1. **Classify by hand.** If you know your study system, group taxa
   yourself. This is the approach used successfully for every real
   dataset this project has analyzed so far -- a simple
   `dplyr::case_when()` over a taxonomic column is usually enough:

```{r sampling-groups, eval = FALSE, purl = FALSE}
# Build the sampling_group column yourself, directly on the taxonomy you
# already have -- no gridding or binning step is needed first, since
# estimate_kernel_priors() reads sampling_group_col straight off
# occurrence_data.
occurrences_grouped <- occurrences |>
  dplyr::mutate(
    sampling_group = dplyr::case_when(
      class == "Actinopteri"   ~ "fishes",       # same eDNA marker/filter
      class %in% c("Aves")     ~ "birds",        # standardized point counts
      phylum == "Arthropoda"   ~ "invertebrates", # kick-net samples
      TRUE                     ~ "other"
    )
  )

table(occurrences_grouped$sampling_group)

kernel_fit_grouped <- estimate_kernel_priors(
  occurrence_data     = occurrences_grouped,
  site_lat            = 34.45,
  site_lon            = -120.47,
  site_habitat        = "Marine",
  lambda_km           = lambda_km,
  sampling_group_col  = "sampling_group"
)
```

2. **Build separate models per survey type**, and pass the appropriate
   priors to TaxaAssign for the relevant taxonomic hypotheses.

With `sampling_group_col` supplied, both the composition estimate and
the dark-diversity budget (Step 4) are computed *within* each group
rather than pooled across all of them. A barely-sampled group (say, a
handful of incidental plant records collected alongside a fish survey)
can then no longer dilute a well-sampled group's estimates, or inflate
the dark-diversity floor for species the assay could never have
detected in the first place.

Don't hold back from splitting a group finely because you're worried
about sample size -- `estimate_kernel_priors()` does not need each
group to individually "have enough data" for statistical adequacy. A
sparse group just gets an honestly wide `theta_sd` instead of a wrong,
falsely-precise one; the estimator has been confirmed to fit cleanly
down to a single-taxon group. Classify at whatever granularity
genuinely reflects distinct detection methods, and let the model's own
uncertainty do the rest.

For a single-taxon-group dataset (a dedicated fish survey, say), this
option is a no-op by construction: there's only one group to have, so
supplying it changes nothing. It only matters once your occurrence pool
genuinely spans more than one detection process.

## Step 6: Assemble the final prior table

Bring every piece together into one table, in the compositional order
TaxaExpect uses throughout: species with real local evidence, dark
diversity, and (if used) named unobserved evidence:

```{r assemble, eval = FALSE, purl = FALSE}
priors <- dplyr::bind_rows(
  kernel_fit$priors,   # resident_observed: species with local evidence
  undetected           # resident_undetected: singleton mirrors + global floor
  # , priors_with_evidence   # if Step 4's named-evidence step was used
)

table(priors$prior_branch)
```

Every row carries a `prior_branch` column (`"resident_observed"`,
`"resident_undetected"`, or -- if you add domestic/food priors, next --
`"transport"`) that `TaxaAssign::join_priors()` reads to decide how a row
should be treated, plus the composite key
`(taxon_name, taxon_name_rank, grid_id, main_habitat)` it joins on. Hand
this table straight to `join_priors()`:

```{r hand-off, eval = FALSE, purl = FALSE}
library(TaxaAssign)

posteriors <- likelihoods_df |>
  join_priors(priors, backbone_id = 4L) |>
  compute_posterior()
```

### Optional: domestic and food species

Occurrence databases like GBIF systematically under-represent captive
and domestic organisms -- pets, livestock, food crops -- because most
citizen-science and museum records describe wild populations. Left
alone, a genuinely plausible detection of, say, a food fish near a
processing facility gets the same tiny dark-diversity floor as a
species with no plausible reason to be there at all.
`generate_domestic_food_priors()` populates a separate `transport`
branch for exactly this case, using fixed lists of known domestic
animals, food species, and cultivars, plus (optionally) a live check
against local iNaturalist records for anything not on those lists:

```{r domestic-food, eval = FALSE, purl = FALSE}
transport_priors <- generate_domestic_food_priors(
  model_obj       = kernel_fit,
  lat             = 34.45,
  lng             = -120.47,
  grid_id         = kernel_fit$params$site_id,
  match_list_taxa = unique(likelihoods_df$taxon_name)   # taxa with real likelihoods this run
)

priors <- dplyr::bind_rows(priors, transport_priors)
```

This is a different kind of claim than the occurrence-plausibility
evidence in Step 4: a domestic/food detection means "this species is
plausible here because of human transport," not "a wild population is
probably present," so the two are never combined into one row.

## Step 7: Visualize the result

`plot_theta_surface()` evaluates the same kernel estimator continuously
across a lattice of points, rather than at one site, so you can see how
a species' prior varies across the whole study area:

```{r visualize, eval = FALSE, purl = FALSE}
plot_theta_surface(
  kernel_fit,
  occurrence_data = occurrences,
  taxon           = "Girella nigricans"
)

# An interactive leaflet map, with a hover label per cell:
plot_theta_surface(
  kernel_fit,
  occurrence_data = occurrences,
  taxon           = c("Girella nigricans", "Semicossyphus pulcher"),
  interactive     = TRUE
)
```

Because it evaluates the identical formula `estimate_kernel_priors()`
uses, the surface's value at the site's own coordinates reproduces
`kernel_fit$priors$theta_mean` exactly. It's a picture of the
estimator, not a separate approximation of it.

## Where to Learn More

This vignette covers the workflow end to end but keeps the statistics
light. For the full derivation (the exact kernel formula, how the
Good-Turing/Chao budget for undetected species is priced, the validation
results on real datasets, and the assumptions and limitations that
apply), see
[`inst/TaxaExpect_supplemental_methods.md`](../inst/TaxaExpect_supplemental_methods.md),
and the function help pages (`?estimate_kernel_priors`,
`?calibrate_kernel_bandwidth`, `?apply_undetected_evidence`) for full
argument documentation.
