---
title: "Building Occurrence-Based Priors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building Occurrence-Based Priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Purpose

TaxaExpect generates prior probabilities for species occurrence at a given
location and habitat. These priors feed into TaxaAssign's Bayesian workflow,
where they are combined with likelihoods from TaxaLikely to produce posterior
taxonomic assignments.

The core question TaxaExpect answers: *"Before seeing any match data, how
likely is it that this species occurs at this site?"*

## Quick Start

```{r quickstart}
library(TaxaExpect)

bp <- build_priors(
  taxa = c("Fundulidae", "Atherinopsidae", "Gobiidae"),
  lat = 34.4, lon = -119.8,
  geographic_context = "Santa Barbara Channel, California"
)

head(bp$priors)
print(bp$model)
```

## Detailed Walkthrough

### Step 1: Fetch GBIF occurrences (TaxaFetch)

```{r fetch}
library(TaxaFetch)

keys <- get_keys_from_context(data.frame(taxon_name = c("Fundulidae", "Gobiidae")))
bbox <- make_bbox_wkt(lat = 34.4, lon = -119.8, radius_deg = 2)

gbif_raw <- fetch_gbif_occurrences(keys$usageKey, bbox = bbox)
occurrences <- filter_gbif_quality(gbif_raw)
```

### Step 2: Assign habitat (TaxaHabitat)

```{r habitat}
library(TaxaHabitat)

prompt <- build_habitat_prompt(
  unique(occurrences$taxon_name),
  geographic_context = "Southern California"
)

# Submit to LLM (one chunk at a time)
raw_text <- llm_fn(prompt$prompts[[1]])
habitats_df <- parse_hierarchical_habitat_response(raw_text, prompt$chunks[[1]], prompt)

occurrences <- assign_habitat_biological(occurrences, habitats_df)
```

### Step 3: Spatial gridding

```{r grid}
grid_result <- optimize_grid_size(occurrences, n_covariates = 2)
occurrences_gridded <- create_sites_from_grid(occurrences, grid_size = grid_result$best_grid)
```

### Step 4: Prepare model data

```{r model-data}
model_data <- prepare_model_dataframe(occurrences_gridded)

basis <- compute_moran_basis(unique(model_data$grid_id), k = 5)
model_data <- dplyr::left_join(model_data, basis, by = "grid_id")
```

### Step 5: Train biodiversity model

```{r train}
model_fit <- screen_spatial_formula(
  data         = model_data,
  formula_full = cbind(n_species, n_other) ~ main_habitat +
    (1 | taxon_name) + diag(main_habitat | taxon_name) +
    (0 + B1 | taxon_name) + (0 + B2 | taxon_name),
  sd_threshold = 0.20
)
```

### Step 6: Generate priors

```{r priors}
priors_observed   <- generate_full_priors(model_fit, new_sites = model_data)
priors_undetected <- generate_undetected_diversity(model_fit)
priors_combined   <- dplyr::bind_rows(priors_observed, priors_undetected)
```

### Step 7: Translate to target backbone

```{r translate}
library(TaxaTools)

ncbi_lookup <- verify_taxon_names(unique(priors_combined$taxon_name), backbone_id = 4) |>
  change_backbone(input_col = "user_supplied_name")

taxaexpect_priors <- priors_combined |>
  dplyr::left_join(ncbi_lookup, by = c("taxon_name" = "gbif_name")) |>
  create_taxon_names(rank_system = c("family", "genus", "species"))
```

## Key Parameters

| Parameter | Default | When to change |
|---|---|---|
| `search_radius_deg` | 2 | Increase for data-poor regions |
| `moran_k` | 5 | Reduce if model fails to converge |
| `sd_threshold` | 0.20 | Lower to keep more spatial terms |
| `target_backbone_id` | 4 (NCBI) | Change to match your reference DB |
| `habitat_scheme` | NULL (3-cat) | Use "IUCN_L1" for finer classification |
| `checkpoint_dir` | NULL | Set a path for crash recovery |

## Common Pitfalls

1. **Zero GBIF records:** Increase `search_radius_deg` or add
   `supplemental_occurrences` from local surveys.

2. **Model convergence failure:** `build_priors()` returns `$model = NULL`
   with a warning. Common with few species or small spatial extent. Try
   reducing `moran_k` or using a simpler formula.

3. **glmmTMB/TMB version mismatch:** Reinstall glmmTMB from source:
   `install.packages("glmmTMB", type = "source")`.

4. **Habitat column mismatch:** TaxaAssign's `join_priors()` joins on
   `main_habitat`. Ensure the habitat names in priors match your site
   specification.
