---
title: "Quality Flagging with TaxaFlag"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quality Flagging with TaxaFlag}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

TaxaFlag provides three independent flagging mechanisms for identifying
anomalous detections in taxonomic assignment results:

1. **Contamination flagging** -- compare taxon proportions against control samples
2. **Handler artifact flagging** -- detect temporal clustering near equipment events
3. **Expert review** -- LLM-based assessment of assignment plausibility

Each flag function appends categorical columns to your data. You decide how to
weight and combine them.

## Contamination Flagging

The `flag_contaminant()` function compares read proportions in your samples
against control samples (lab blanks, field blanks, or positive controls).

```{r contamination-example, eval = FALSE}
library(TaxaFlag)

# Identify control samples
controls <- c("blank_1", "blank_2", "blank_3")

# Flag taxa that appear disproportionately in blanks
flagged <- flag_contaminant(
  input_df               = reads_long,
  control_samples  = controls,
  taxon_col        = "taxon_name",
  reads_col        = "n_reads",
  event_col        = "event_id",
  contaminant_type = "lab_contaminant"
)

# Output adds: observation_validity, validity_flag, validity_reason
table(flagged$validity_flag)
```

`validity_flag` values (the qualifier reflects `contaminant_type`, here
`"lab_contaminant"`):
- `"valid"` -- taxon found mainly in field samples (likely genuine detection)
- `"questionable_lab_contaminant"` -- elevated in controls but not conclusive
- `"invalid_lab_contaminant"` -- strong evidence of contamination (found mainly in controls)

`observation_validity` is the numeric score behind the flag -- high = good/genuine,
low = likely contaminant.

## Handler Artifact Flagging

The `flag_handler()` function identifies detections clustered near equipment
setup, collection, or handling events.

By default, "deployment edges" are the detection data's own earliest/latest
timestamp per station -- **not** the true setup/pickup time, so the very
first and last genuine wildlife detection at a station is always scored
maximally suspect. If you have a real per-station deployment log, pass it as
`station_metadata` to anchor edges on the actual setup/pickup times instead:

```{r handler-example, eval = FALSE}
# Flag detections within 30 minutes of sample collection
flagged <- flag_handler(
  input_df               = assignments,
  datetime_col     = "detection_time",
  group_col        = "station_id",
  interval_minutes = 30
)

# Result adds: observation_validity, validity_flag, validity_reason
# validity_flag values: "valid" / "questionable_handling" / "invalid_handling"
table(flagged$validity_flag)

# Better: anchor edges on real deploy/retrieve times, not the data's own
# min/max, so a genuine detection right when the camera happened to start
# recording isn't automatically flagged as a handler artifact.
# station_deploy_log needs one row per station_id with real timestamps:
#   station_id, deploy_time, retrieve_time
flagged_anchored <- flag_handler(
  input_df               = assignments,
  datetime_col     = "detection_time",
  group_col        = "station_id",
  interval_minutes = 30,
  station_metadata = station_deploy_log
)

# edge_anchor_source tells you, per row, whether the real deployment log or
# the data-derived fallback was used for that station
table(flagged_anchored$edge_anchor_source)
```

## Expert Review (LLM-based)

The `review_assignments()` function submits assignments to an LLM for
structured expert review across multiple dimensions.

```{r review-example, eval = FALSE}
# Requires an LLM provider function and a context list
reviewed <- review_assignments(
  input_df        = consensus_df,
  taxon_col = "consensus_taxon",
  context   = list(geography = "Southern California Bight",
                   habitat   = "Marine"),
  target_group = "fish",
  llm_fn    = TaxaTools::call_api
)

# Adds 8 structured columns (llm_ prefixed since 2026-09-06 -- these are
# independent LLM judgments, never derived from the pipeline's own values):
# llm_habitat_plausibility, llm_geographic_plausibility, llm_scope_plausibility,
# llm_contamination_risk, review_alternatives, review_lower_hypotheses,
# review_confidence, review_comment
```

Plausibility columns (`llm_habitat_plausibility`, `llm_geographic_plausibility`,
`llm_scope_plausibility`) use `"likely"` / `"possible"` / `"unlikely"` — higher
means more plausible as a genuine detection.

`llm_contamination_risk` uses `"low"` / `"moderate"` / `"high"` — higher means
more concern about contamination.

## Combining Flags

TaxaFlag intentionally does not provide a `combine_flags()` function. The
appropriate weighting depends on your study design and risk tolerance:

```{r combine-example, eval = FALSE}
# Example: conservative approach (flag if ANY mechanism triggers)
# lab_flags/handler_flags are each the output of a separate flag_contaminant()/
# flag_handler() call joined onto the same taxon/observation, so their
# validity_flag columns need a per-check prefix -- see the README's
# "Combine results" example for a full join.
flagged$any_flag <- with(flagged,
  lab_validity_flag    == "invalid_lab_contaminant" |
  handler_validity_flag == "invalid_handling" |
  llm_contamination_risk %in% c("high", "moderate") |
  review_confidence    == "low"
)

# Example: permissive approach (only flag if multiple mechanisms agree)
flagged$multi_flag <- with(flagged,
  (lab_validity_flag == "invalid_lab_contaminant") +
  (llm_contamination_risk %in% c("high", "moderate")) >= 2
)
```

## Reporting

Generate a structured report section for Methods/Results:

```{r report-example, eval = FALSE}
sec <- report_flags(flagged)
print(sec)

# Or assemble with other TaxaID package reports:
full_report <- TaxaTools::assemble_report(fetch_sec, match_sec, assign_sec, sec)
```
