---
title: "Quality Flagging with TaxaFlag"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quality Flagging with TaxaFlag}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

TaxaFlag provides three independent flagging mechanisms for identifying
anomalous detections in taxonomic assignment results:

1. **Contamination flagging** -- compare taxon proportions against control samples
2. **Handler artifact flagging** -- detect temporal clustering near equipment events
3. **Expert review** -- LLM-based assessment of assignment plausibility

Each flag function appends categorical columns to your data. You decide how to
weight and combine them.

## Contamination Flagging

The `flag_contaminant()` function compares read proportions in your samples
against control samples (lab blanks, field blanks, or positive controls).

```{r contamination-example, eval = FALSE}
library(TaxaFlag)

# Identify control samples
controls <- c("blank_1", "blank_2", "blank_3")

# Flag taxa that appear disproportionately in blanks
flagged <- flag_contaminant(
  df               = reads_long,
  control_samples  = controls,
  taxon_col        = "taxon_name",
  reads_col        = "n_reads",
  event_col        = "event_id",
  contaminant_type = "lab_contaminant"
)

# Output adds: lab_contaminant_risk, lab_contaminant_score, lab_contaminant_reason
table(flagged$lab_contaminant_risk)
```

Risk values:
- `"low"` -- taxon found mainly in field samples (likely genuine detection)
- `"moderate"` -- elevated in controls but not conclusive
- `"high"` -- strong evidence of contamination (found mainly in controls)

## Handler Artifact Flagging

The `flag_handler()` function identifies detections clustered near equipment
setup, collection, or handling events.

```{r handler-example, eval = FALSE}
# Flag detections within 30 minutes of sample collection
flagged <- flag_handler(
  df               = assignments,
  datetime_col     = "detection_time",
  group_col        = "station_id",
  interval_minutes = 30
)

# Result adds: flag_handler column
table(flagged$flag_handler)
```

## Expert Review (LLM-based)

The `review_assignments()` function submits assignments to an LLM for
structured expert review across multiple dimensions.

```{r review-example, eval = FALSE}
# Requires an LLM provider function and a context list
reviewed <- review_assignments(
  df        = consensus_df,
  taxon_col = "consensus_taxon",
  context   = list(geography = "Southern California Bight",
                   habitat   = "Marine"),
  target_group = "fish",
  llm_fn    = TaxaTools::call_api
)

# Adds 8 structured columns:
# habitat_plausibility, geographic_plausibility, scope_plausibility, contamination_risk,
# review_alternatives, review_lower_hypotheses, review_confidence, review_comment
```

Plausibility columns (`habitat_plausibility`, `geographic_plausibility`,
`scope_plausibility`) use `"likely"` / `"possible"` / `"unlikely"` — higher
means more plausible as a genuine detection.

`contamination_risk` uses `"low"` / `"moderate"` / `"high"` — higher means
more concern about contamination.

## Combining Flags

TaxaFlag intentionally does not provide a `combine_flags()` function. The
appropriate weighting depends on your study design and risk tolerance:

```{r combine-example, eval = FALSE}
# Example: conservative approach (flag if ANY mechanism triggers)
flagged$any_flag <- with(flagged,
  lab_contaminant_risk == "high" |
  flag_handler         == "unlikely" |
  contamination_risk   %in% c("high", "moderate") |
  review_confidence    == "low"
)

# Example: permissive approach (only flag if multiple mechanisms agree)
flagged$multi_flag <- with(flagged,
  (lab_contaminant_risk == "high") +
  (contamination_risk %in% c("high", "moderate")) >= 2
)
```

## Reporting

Generate a structured report section for Methods/Results:

```{r report-example, eval = FALSE}
sec <- report_flags(flagged)
print(sec)

# Or assemble with other TaxaID package reports:
full_report <- TaxaTools::assemble_report(fetch_sec, match_sec, assign_sec, sec)
```
