---
title: "Assigning Habitat via LLM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Assigning Habitat via LLM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, eval = FALSE, purl = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  # purl = FALSE is load-bearing, not redundant with eval = FALSE.
  # R CMD check's "running R code from vignettes" step TANGLES the vignette
  # (knitr::purl()) and sources the result -- purl() honours purl=, not eval=,
  # so without this a documentation-only vignette still executes during check
  # and errors on objects its prose never defines.
  purl = FALSE
)
```

<!--
  EVERY chunk in this vignette carries `eval = FALSE, purl = FALSE` in its own
  header, and that repetition is load-bearing -- do not DRY it up
  into the setup chunk. R CMD check's "running R code from vignettes" step
  TANGLES the vignette with knitr::purl(), which does NOT execute the setup
  chunk, so an opts_chunk$set(purl = FALSE) there has no effect at tangle time
  and every chunk is sourced. Same fix as TaxaExpect/vignettes/building-priors.Rmd.
-->

## Purpose

TaxaHabitat assigns habitat types to species and occurrence records using LLM
(Large Language Model) inference. It produces per-species habitat weight
vectors and per-point habitat classifications that feed into TaxaExpect for
biodiversity modelling.

Every code chunk below is illustrative and is not executed when the package
is checked, since it calls live LLM services; a test in this package's own
suite verifies every call and argument name against the installed package
instead.

## Quick Start

```{r quickstart, eval = FALSE, purl = FALSE}
library(TaxaHabitat)
library(TaxaTools)

# Build and submit habitat prompt
prompt <- build_habitat_prompt(
  c("Fundulus parvipinnis", "Atherinops affinis", "Mugil cephalus"),
  geographic_context = "Southern California"
)

raw_text <- call_anthropic_api(prompt$prompts[[1]])
habitats <- parse_hierarchical_habitat_response(raw_text, prompt$chunks[[1]], prompt)

# Assign habitat to occurrence points
occurrences <- assign_habitat_biological(occurrence_data, habitats)
```

## Detailed Walkthrough

### Step 1: Choose a habitat scheme

```{r scheme, eval = FALSE, purl = FALSE}
# Default: 3-category (Marine / Freshwater / Terrestrial)
prompt <- build_habitat_prompt(species_list, habitat_scheme = NULL)

# IUCN Level 1 (16 categories)
prompt <- build_habitat_prompt(species_list, habitat_scheme = "IUCN_L1")

# Custom scheme
my_scheme <- data.frame(
  habitat_name = c("Rocky Intertidal", "Sandy Beach", "Kelp Forest", "Estuary"),
  stringsAsFactors = FALSE
)
prompt <- build_habitat_prompt(species_list, habitat_scheme = my_scheme)
```

### Step 2: Submit to LLM

The prompt is automatically chunked for large species lists.

```{r submit, eval = FALSE, purl = FALSE}
raw_texts <- character(prompt$n_chunks)
for (i in seq_len(prompt$n_chunks)) {
  raw_texts[i] <- call_anthropic_api(prompt$prompts[[i]])
}

# Parse all chunks
habitats_list <- lapply(seq_len(prompt$n_chunks), function(i) {
  parse_hierarchical_habitat_response(raw_texts[i], prompt$chunks[[i]], prompt)
})
habitats_df <- do.call(rbind, habitats_list)
```

### Step 3: Assign habitat to points

```{r assign, eval = FALSE, purl = FALSE}
occurrences <- assign_habitat_biological(
  occurrence_data = occurrence_data,
  habitats_df = habitats_df,
  threshold = 0.3
)

# Check consensus
table(occurrences$main_habitat, useNA = "always")
```

### Step 4: Flag spatial inconsistencies

```{r flag, eval = FALSE, purl = FALSE}
occurrences <- flag_habitat_inconsistencies(occurrences)
# Review flagged points
table(occurrences$spatial_flag)
```

## Key Parameters

| Parameter | Default | When to change |
|---|---|---|
| `habitat_scheme` | NULL (3-cat) | Use "IUCN_L1" or custom for finer resolution |
| `geographic_context` | NULL | Always supply for better ecoregion inference |
| `threshold` | 0.3 | Lower for transitional habitats (estuaries) |
| `chunk_size` | 60 | Reduce if LLM truncates responses |

## Common Pitfalls

1. **LLM returns unexpected columns:** `parse_hierarchical_habitat_response()`
   is tolerant of extra columns but requires the scheme's habitat columns.
   Check `prompt$habitat_cols` against the parsed output.

2. **All habitats "Other":** The species may not fit the scheme. Check
   `habitats_df$habitat_best_guess` for the LLM's actual suggestion, and
   consider a finer scheme.

3. **Spatial flags too aggressive:** Coastal species near shore boundaries
   are often flagged. Increase `coast_buffer_m` in
   `flag_habitat_inconsistencies()`.
