---
title: "Assigning Habitat via LLM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Assigning Habitat via LLM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Purpose

TaxaHabitat assigns habitat types to species and occurrence records using LLM
(Large Language Model) inference. It produces per-species habitat weight
vectors and per-point habitat classifications that feed into TaxaExpect for
biodiversity modelling.

## Quick Start

```{r quickstart}
library(TaxaHabitat)
library(TaxaTools)

# Build and submit habitat prompt
prompt <- build_habitat_prompt(
  c("Fundulus parvipinnis", "Atherinops affinis", "Mugil cephalus"),
  geographic_context = "Southern California"
)

raw_text <- call_anthropic_api(prompt$prompts[[1]])
habitats <- parse_hierarchical_habitat_response(raw_text, prompt$chunks[[1]], prompt)

# Assign habitat to occurrence points
occurrences <- assign_habitat_biological(occurrence_data, habitats)
```

## Detailed Walkthrough

### Step 1: Choose a habitat scheme

```{r scheme}
# Default: 3-category (Marine / Freshwater / Terrestrial)
prompt <- build_habitat_prompt(species_list, habitat_scheme = NULL)

# IUCN Level 1 (16 categories)
prompt <- build_habitat_prompt(species_list, habitat_scheme = "IUCN_L1")

# Custom scheme
my_scheme <- data.frame(
  habitat_name = c("Rocky Intertidal", "Sandy Beach", "Kelp Forest", "Estuary"),
  stringsAsFactors = FALSE
)
prompt <- build_habitat_prompt(species_list, habitat_scheme = my_scheme)
```

### Step 2: Submit to LLM

The prompt is automatically chunked for large species lists.

```{r submit}
raw_texts <- character(prompt$n_chunks)
for (i in seq_len(prompt$n_chunks)) {
  raw_texts[i] <- call_anthropic_api(prompt$prompts[[i]])
}

# Parse all chunks
habitats_list <- lapply(seq_len(prompt$n_chunks), function(i) {
  parse_hierarchical_habitat_response(raw_texts[i], prompt$chunks[[i]], prompt)
})
habitats_df <- do.call(rbind, habitats_list)
```

### Step 3: Assign habitat to points

```{r assign}
occurrences <- assign_habitat_biological(
  data        = occurrence_data,
  habitats_df = habitats_df,
  threshold   = 0.3
)

# Check consensus
table(occurrences$main_habitat, useNA = "always")
```

### Step 4: Flag spatial inconsistencies

```{r flag}
occurrences <- flag_habitat_inconsistencies(occurrences)
# Review flagged points
table(occurrences$spatial_flag)
```

## Key Parameters

| Parameter | Default | When to change |
|---|---|---|
| `habitat_scheme` | NULL (3-cat) | Use "IUCN_L1" or custom for finer resolution |
| `geographic_context` | NULL | Always supply for better ecoregion inference |
| `threshold` | 0.3 | Lower for transitional habitats (estuaries) |
| `chunk_size` | 60 | Reduce if LLM truncates responses |

## Common Pitfalls

1. **LLM returns unexpected columns:** `parse_hierarchical_habitat_response()`
   is tolerant of extra columns but requires the scheme's habitat columns.
   Check `prompt$habitat_cols` against the parsed output.

2. **All habitats "Other":** The species may not fit the scheme. Check
   `habitats_df$habitat_best_guess` for the LLM's actual suggestion, and
   consider a finer scheme.

3. **Spatial flags too aggressive:** Coastal species near shore boundaries
   are often flagged. Increase `coast_buffer_m` in
   `flag_habitat_inconsistencies()`.
