---
title: "TaxaID Ecosystem: End-to-End Taxonomic Assignment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TaxaID Ecosystem: End-to-End Taxonomic Assignment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

The TaxaID ecosystem is a suite of eight R packages that implements Bayesian
taxonomic assignment for biological observations -- eDNA sequences, camera trap
images, acoustic recordings, or any data where an observation must be matched to
a taxon. Given a set of candidate matches (with scores), TaxaID converts those
scores into likelihoods, estimates priors from occurrence data, computes
posterior probabilities via Bayes' theorem, and flags anomalous detections.

### Pipeline Diagram

```
Raw observations
       |
   TaxaMatch ........... standardize match data
       |
       v
   TaxaLikely .......... score -> likelihood (hierarchical Bayesian model)
       |
       |          TaxaTools .......... name cleaning, verification, utilities
       |              |
       |          TaxaFetch .......... fetch GBIF / DataONE occurrences
       |              |
       |          TaxaHabitat ........ assign habitat via LLM
       |              |
       |          TaxaExpect ......... fit spatial model -> priors
       |              |
       v              v
   TaxaAssign .................. posterior = likelihood x prior
       |
       v
   TaxaFlag ................... flag anomalous detections
       |
       v
   Consensus taxonomy (species, genus, family, etc.)
```

The left branch (TaxaMatch -> TaxaLikely) handles **likelihoods**: how well does
the observation match each candidate taxon? The right branch (TaxaFetch ->
TaxaHabitat -> TaxaExpect) handles **priors**: how likely is each taxon to occur
at this location? TaxaAssign combines them.

### Two Pathways

TaxaID offers two workflows for taxonomic assignment:

| | Bayesian Pathway | LLM Pathway |
|---|---|---|
| **Priors from** | Occurrence data (GBIF) + spatial model | LLM expert knowledge |
| **Likelihoods from** | Trained hierarchical model (TaxaLikely) | Trained model or LLM scores |
| **Best for** | Large datasets, reproducible pipelines | Quick exploration, data-poor regions |
| **Packages used** | All 8 | TaxaMatch + TaxaAssign + TaxaFlag (+ TaxaTools) |
| **API calls** | GBIF + NCBI | LLM provider (Anthropic, OpenAI, etc.) |

Both pathways end at `TaxaAssign::posterior_consensus()` and can be compared
with `TaxaAssign::score_consensus()`.

---

## Bayesian Pathway: Full Pipeline

### Step 1: Standardize match data (TaxaMatch)

```{r match}
library(TaxaMatch)

# Read DADA2 sequence table
seqs <- read_sequence_table(
  seqtab_path = "seqtab_nochim.rds",
  taxonomy_path = "tax_table.rds"
)

# Filter by length and abundance
seqs_filtered <- filter_sequences(seqs, barcode_term = "MiFish")

# BLAST against NCBI
blast_results <- blast_sequences(
  seqs_filtered,
  barcode_term = "MiFish",
  score_range = 5
)

# Standardize to canonical match object
match_df <- standardize_match_data(
  blast_results,
  rank_system = c("family", "genus", "species")
)

# Remove higher-rank duplicates
match_df <- filter_redundant_hypotheses(match_df)
```

### Step 2: Build reference and train likelihood model (TaxaLikely)

```{r likelihood}
library(TaxaLikely)

# Fetch reference sequences from NCBI
reference_df <- fetch_reference_sequences(
  taxa = unique(match_df$family),
  barcode_term = "MiFish",
  max_per_species = 5
)

# Build alignment matrix
ref_matrix <- build_reference_matrix(reference_df)

# Flag mislabeled references
errors <- flag_reference_errors(ref_matrix)

# Train the hierarchical model
model <- train_likelihood_model(ref_matrix, rank_system = c("family", "genus", "species"))

# Convert match scores to likelihoods
result <- evaluate_likelihoods(match_df, model)
likelihoods <- result$likelihoods

# Keep finest-rank candidates
likelihoods <- filter_top_hypotheses(likelihoods)
```

### Step 3: Fetch occurrence data (TaxaFetch)

```{r fetch}
library(TaxaFetch)

keys <- get_keys_from_context(
  data.frame(taxon_name = unique(match_df$family))
)
bbox <- make_bbox_wkt(lat = 34.4, lon = -119.8, radius_deg = 2)
gbif_raw <- fetch_gbif_occurrences(keys$usageKey, bbox = bbox)
occurrences <- filter_gbif_quality(gbif_raw)
```

### Step 4: Assign habitat (TaxaHabitat)

```{r habitat}
library(TaxaHabitat)

prompt <- build_habitat_prompt(
  unique(occurrences$species),
  geographic_context = "Southern California"
)
raw_text <- TaxaTools::call_anthropic_api(prompt$prompts[[1]])
habitats <- parse_hierarchical_habitat_response(raw_text, prompt$chunks[[1]], prompt)
occurrences <- assign_habitat_biological(occurrences, habitats)
```

### Step 5: Build priors (TaxaExpect)

```{r priors}
library(TaxaExpect)

grid <- create_sites_from_grid(occurrences, grid_size = 0.25)
model_fit <- fit_occupancy_model(grid$data)
priors <- generate_full_priors(model_fit, grid)
```

### Step 5 (alternative): One-call wrapper

```{r build_priors}
# build_priors() wraps Steps 3-5 into a single call
prior_result <- TaxaExpect::build_priors(
  taxa = unique(match_df$family),
  lat = 34.4, lon = -119.8,
  llm_fn = TaxaTools::call_anthropic_api,
  verbose = TRUE
)
priors <- prior_result$priors
```

### Step 6: Compute posteriors (TaxaAssign)

```{r posterior}
library(TaxaAssign)

# Detect unreferenced species
unreferenced <- suggest_unreferenced_species(
  match_df,
  llm_fn = TaxaTools::call_anthropic_api,
  barcode_term = "MiFish"
)

# Expand generic unreferenced rows into named species
likelihoods_expanded <- expand_unreferenced_hypotheses(
  likelihoods, unreferenced, priors
)

# Apply coverage constraints (suppress unreferenced for complete genera)
coverage <- audit_barcode_coverage(
  match_df, barcode_term = "MiFish"
)
likelihoods_final <- apply_coverage_constraints(likelihoods_expanded, coverage)

# Join priors
joined <- join_priors(likelihoods_final, priors, site = list(main_habitat = "Marine"))

# Compute posterior
result <- compute_posterior(joined, n_sims = 1000)

# Consensus taxonomy
consensus <- posterior_consensus(result, cumulative_threshold = 0.95)

# Empirical Bayes refinement
result_updated <- update_prior_from_consensus(result, consensus)
consensus_final <- posterior_consensus(result_updated, cumulative_threshold = 0.95)
```

### Step 6 (alternative): One-call wrapper

```{r run_bayesian}
# run_bayesian_pipeline() wraps Step 6 into a single call
pipeline <- run_bayesian_pipeline(
  match_df = match_df,
  model_params = model,
  taxaexpect_priors = priors,
  site = list(main_habitat = "Marine"),
  barcode_term = "MiFish",
  verbose = TRUE
)
consensus_final <- pipeline$consensus
```

---

## LLM Pathway: Quick Assignment

The LLM pathway skips the occurrence data pipeline entirely. An LLM assigns
priors based on expert knowledge of species ranges and habitat.

```{r llm_manual}
library(TaxaAssign)
library(TaxaTools)

# Build context (auto-populates habitat, ecoregion, date via LLM)
ctx <- build_context(
  match_df,
  llm_fn = call_anthropic_api,
  geographic_hint = "Southern California estuaries"
)

# LLM assigns priors + detects unreferenced species in one call
result <- assign_taxa_llm(
  match_df, ctx,
  llm_fn = call_anthropic_api,
  detect_unreferenced = TRUE,
  barcode_term = "MiFish"
)

# Posterior computation and consensus
posterior <- compute_posterior(result, n_sims = 1000)
consensus <- posterior_consensus(posterior, cumulative_threshold = 0.95)
result_updated <- update_prior_from_consensus(posterior, consensus)
consensus_final <- posterior_consensus(result_updated, cumulative_threshold = 0.95)
```

### LLM one-call wrapper

```{r run_llm}
pipeline <- run_llm_pipeline(
  match_df = match_df,
  llm_fn = call_anthropic_api,
  geographic_hint = "Southern California estuaries",
  detect_unreferenced = TRUE,
  barcode_term = "MiFish",
  generate_report = TRUE,
  verbose = TRUE
)
consensus_final <- pipeline$consensus
report <- pipeline$report
```

---

## Comparing Approaches

Both pathways produce posterior-based consensus. You can also run a conventional
score-based consensus for comparison:

```{r compare}
# Score-based consensus (no model needed)
score_cons <- score_consensus(match_df)

# Compare
comparison <- merge(
  consensus_final[, c("observation_id", "consensus_taxon", "consensus_rank")],
  score_cons[, c("observation_id", "consensus_taxon", "consensus_rank")],
  by = "observation_id", suffixes = c("_posterior", "_score")
)

# Where do they disagree?
disagree <- comparison[comparison$consensus_taxon_posterior != comparison$consensus_taxon_score, ]
```

### When to Use Which

**Use the Bayesian pathway when:**

- You have a well-characterized reference database
- Occurrence data are available for your study region
- Reproducibility is critical (no LLM stochasticity)
- You are processing large datasets (model trains once, applies to all)

**Use the LLM pathway when:**

- You are exploring a new dataset quickly
- Occurrence data are sparse or unavailable
- You want expert-level context without building a spatial model
- You are working in a data-poor region where GBIF coverage is thin

**Use score consensus when:**

- You want a quick baseline with no modeling
- You need to compare against the traditional approach
- Your match scores are highly reliable (e.g., 99%+ identity)

---

## Generating a Report

Both pathways support automated report generation:

```{r report}
report <- generate_report(
  result = result_updated,
  consensus = consensus_final,
  unreferenced_result = unreferenced,
  data_type = "eDNA",
  marker = "MiFish 12S",
  study_description = "Estuarine fish biodiversity survey in Southern California",
  llm_fn = call_anthropic_api
)

cat(report$methods)
cat(report$results)
```

---

## Quality Flagging (TaxaFlag)

After assignment, screen results for anomalous detections:

```{r flag}
library(TaxaFlag)

# Flag potential contaminants from lab blanks
flagged <- flag_contaminant(
  consensus_final, control_samples = c("blank_1", "blank_2"),
  taxon_col = "consensus_taxon", reads_col = "n_reads",
  event_col = "event_id", contaminant_type = "lab"
)

# LLM expert review of assignment plausibility
reviewed <- review_assignments(
  flagged, taxon_col = "consensus_taxon",
  site_col = "site_name", llm_fn = call_anthropic_api
)
```

---

## Assembled Reports

Each package provides a `report_*()` function that captures what it did.
These can be assembled into a unified Methods + Results document:
```{r assemble}
library(TaxaTools)

sections <- list(
  TaxaMatch::report_match(match_df),
  report_likelihood(model),
  report_assign(result_updated, consensus_final),
  TaxaFlag::report_flags(reviewed)
)

full_report <- assemble_report(sections)
cat(format(full_report))
```

---

## Key Concepts

### Hypothesis Types

Every candidate match generates up to three hypothesis types:

- **specific_candidate**: a taxon present in the reference database (H1)
- **unreferenced_species**: a described species absent from the reference (H2)
- **unreferenced_genus**: an undescribed or unsampled genus-level taxon (H3)

The likelihood model assigns different score distributions to each type.
H2 scores are expected to be lower than H1 (sister species, not exact match);
H3 scores lower still.

### Prior Sources

- **TaxaExpect priors**: derived from GBIF occurrence data + spatial occupancy model.
  Quantitative, reproducible, but limited by GBIF coverage.
- **LLM priors**: assigned by an LLM based on species range, habitat, and
  geographic context. Flexible, but stochastic and non-reproducible.
- **Flat priors**: equal probability for all candidates. Used as fallback when
  no prior information is available.

### Consensus Methods

- **posterior_consensus()**: Bayesian LCA (lowest common ancestor) among
  hypotheses exceeding a cumulative posterior threshold. Accounts for
  uncertainty.
- **score_consensus()**: traditional rank thresholds (e.g., species at 98%,
  genus at 95%). Deterministic, no model needed.
- **update_prior_from_consensus()**: empirical Bayes refinement -- confirmed
  species from resolved samples boost priors in unresolved samples.

---

## Package Versions and Installation

```{r install}
# Install all TaxaID packages from local source
pkg_paths <- c(
  "TaxaTools", "TaxaFetch", "TaxaHabitat",
  "TaxaMatch", "TaxaLikely", "TaxaExpect", "TaxaAssign", "TaxaFlag"
)
for (pkg in pkg_paths) {
  devtools::install(file.path("path/to/TaxaID", pkg))
}

# Bioconductor dependencies (for TaxaLikely reference matrix building)
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install(c("DECIPHER", "Biostrings"))
```
