---
title: "Bayesian Taxonomic Assignment with TaxaAssign"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Taxonomic Assignment with TaxaAssign}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Purpose

TaxaAssign computes posterior probabilities for taxonomic assignments by
combining likelihoods (from TaxaLikely or LLM-based scores) with priors (from
TaxaExpect or LLM biogeographic knowledge). It provides two parallel workflows:

1. **Bayesian pipeline** (`run_bayesian_pipeline()`): Full statistical model
   using trained likelihoods from TaxaLikely and occurrence-based priors from
   TaxaExpect. Best for publication-quality results.

2. **LLM pipeline** (`run_llm_pipeline()`): Fast approximation using
   score-weighted likelihoods and LLM-estimated priors. Best for exploratory
   analysis with well-known taxa.

Both workflows produce posterior probabilities, consensus taxonomy, and
optional empirical Bayes refinement.

## Quick Start: Bayesian Pipeline

```{r quickstart-bayes}
library(TaxaAssign)

out <- run_bayesian_pipeline(
  match_df          = match_obj,
  model_params      = trained_model,
  taxaexpect_priors = priors,
  site = list(grid_id = "Grid_34p1_m119p1", main_habitat = "Estuarine Bay")
)

head(out$consensus)
```

## Quick Start: LLM Pipeline

```{r quickstart-llm}
out <- run_llm_pipeline(
  match_df        = match_obj,
  geographic_hint = "Southern California",
  barcode_term    = "12S"
)

head(out$consensus)
```

## Detailed Walkthrough: Step-by-Step Bayesian

### Step 1: Evaluate likelihoods (TaxaLikely)

```{r step1}
library(TaxaLikely)

lik_result <- evaluate_likelihoods(match_df, model_params = trained_model)
top_lik    <- filter_top_hypotheses(lik_result$likelihoods)
```

### Step 2: Handle unreferenced species

Unreferenced species — described taxa with no barcode reference — can never
appear as named matches. Expand generic H2/H3 rows into named species.

```{r step2}
# Detect unreferenced species from coverage audit
coverage <- TaxaLikely::audit_barcode_coverage(
  match_df, barcode_term = "12S", target_rank = "genus"
)

# Build unreferenced data frame from TaxaExpect priors
unreferenced_df <- taxaexpect_priors |>
  dplyr::filter(taxon_name %in% coverage$unreferenced) |>
  dplyr::rename(species = taxon_name)

# Expand generic rows into named species
expanded <- expand_unreferenced_hypotheses(top_lik, unreferenced_df)

# Apply coverage constraints (suppress for fully-sampled genera)
constrained <- TaxaLikely::apply_coverage_constraints(expanded, coverage$census)
```

### Step 3: Join priors

```{r step3}
ready <- join_priors(
  likelihoods       = constrained,
  taxaexpect_priors = taxaexpect_priors,
  site              = list(grid_id = "Grid_34p1_m119p1",
                           main_habitat = "Estuarine Bay")
)
```

### Step 4: Compute posteriors

```{r step4}
result <- compute_posterior(ready, n_sims = 1000)
```

### Step 5: Consensus + empirical Bayes refinement

```{r step5}
consensus <- posterior_consensus(result, cumulative_threshold = 0.90)

result_updated <- update_prior_from_consensus(
  result, consensus, presence_multiplier = 5
)

consensus_final <- posterior_consensus(result_updated, cumulative_threshold = 0.90)
```

## Detailed Walkthrough: LLM Pipeline

```{r llm-detail}
# Auto-build context from taxon names
ctx <- build_context(
  taxon_names     = unique(match_df$taxon_name[match_df$score >= 95]),
  geographic_hint = "Southern California",
  date            = "2025"
)

# Detect unreferenced species via LLM
unreferenced <- suggest_unreferenced_species(
  match_df, context = ctx, barcode_term = "12S",
  expand_to_family = TRUE
)

# Run LLM assignment
result <- assign_taxa_llm(
  match_df, context = ctx,
  unreferenced_taxa = unreferenced,
  score_threshold = 80, n_sims = 1000
)

# Consensus + refinement
consensus <- posterior_consensus(result, cumulative_threshold = 0.90)
result_updated <- update_prior_from_consensus(result, consensus)
consensus_final <- posterior_consensus(result_updated)
```

## Key Parameters

| Parameter | Default | When to change |
|---|---|---|
| `cumulative_threshold` | 0.90 | Lower for more liberal assignments |
| `min_posterior` | 0.05 | Raise to exclude weak hypotheses |
| `presence_multiplier` | 5 | Higher = stronger EB boost for confirmed species |
| `n_sims` | 1000 | 0 for point estimates only |
| `score_sharpness` | 0.1 | LLM only: higher makes scores dominate |
| `prior_phi` | c(high=50, moderate=10, low=3) | LLM only: controls prior tightness |

## Comparing Approaches

Use `score_consensus()` for a conventional (non-Bayesian) comparison:

```{r compare}
score_con <- score_consensus(
  match_df, min_score = 80, max_gap = 2,
  rank_thresholds = c(species = 98, genus = 95, family = 90)
)

# Compare
comparison <- dplyr::left_join(
  consensus_final[, c("observation_id", "consensus_taxon", "consensus_rank")],
  score_con[, c("observation_id", "consensus_taxon", "consensus_rank")],
  by = "observation_id", suffix = c("_bayesian", "_score")
)
```

## Common Pitfalls

1. **All posteriors near-uniform:** Priors may be uninformative. Check that
   `join_priors()` found matching priors (look for dark-diversity fallback
   warnings).

2. **"unknown_species" wins:** The LLM pipeline includes a catch-all
   hypothesis. If it dominates, the true species may not be in the candidate
   list. Check `score_threshold` and `unreferenced_taxa`.

3. **TaxaMatch not installed:** `join_priors()` uses
   `TaxaMatch::filter_redundant_hypotheses()` for cleanup. Install TaxaMatch
   or suppress the warning.

4. **NCBI lookup fails in consensus:** `posterior_consensus()` with
   `lookup_missing_taxonomy = TRUE` makes NCBI API calls. Set to `FALSE` for
   offline use.
