---
title: "Fetching Occurrence Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fetching Occurrence Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Purpose

TaxaFetch acquires occurrence data from GBIF, DataONE, BioTIME, PDFs, and
literature searches. It produces standardised occurrence records that feed
into TaxaHabitat (for habitat assignment) and TaxaExpect (for prior
construction).

## Quick Start: GBIF

```{r quickstart}
library(TaxaFetch)

keys <- get_keys_from_context(data.frame(taxon_name = c("Fundulidae", "Gobiidae")))
bbox <- make_bbox_wkt(lat = 34.4, lon = -119.8, radius_deg = 2)

gbif_raw <- fetch_gbif_occurrences(keys$usageKey, bbox = bbox, year_range = "2015,2025")
occurrences <- filter_gbif_quality(gbif_raw)
```

## Data Sources

### GBIF occurrences

```{r gbif}
# Step 1: Get GBIF taxon keys
keys <- get_keys_from_context(
  data.frame(taxon_name = c("Fundulidae", "Atherinopsidae"))
)

# Step 2: Define bounding box
bbox <- make_bbox_wkt(lat = 34.4, lon = -119.8, radius_deg = 2)

# Step 3: Fetch and filter
gbif_raw <- fetch_gbif_occurrences(keys$usageKey, bbox = bbox, limit = 10000)
occurrences <- filter_gbif_quality(gbif_raw, max_coord_uncertainty = 1000)
```

### Combining multiple sources

```{r stack}
# Stack GBIF with local survey data
combined <- stack_occurrences(gbif_occurrences, local_survey_data)
```

### DataONE

```{r dataone}
catalog <- harvest_dataone_catalog(query = "kelp forest fish abundance")
previewed <- preview_dataone_occurrences(catalog[1, ])
occurrences <- fetch_dataone_occurrences(catalog[1, ])
```

### Literature search

```{r literature}
results <- search_literature(
  taxa    = c("Fundulus parvipinnis", "Atherinops affinis"),
  region  = "Southern California"
)
```

## Key Parameters

| Parameter | Default | When to change |
|---|---|---|
| `radius_deg` | 2 | Increase for data-poor regions (~220 km at equator) |
| `year_range` | NULL | Filter to recent years: `"2015,2025"` |
| `limit` | 10000 | Increase for comprehensive coverage |
| `max_coord_uncertainty` | 1000 | Decrease for higher precision |

## Common Pitfalls

1. **No GBIF records returned:** Check that taxon names match GBIF taxonomy.
   Use `get_keys_from_context()` to verify keys before fetching.

2. **Rate limiting:** GBIF and DataONE APIs have rate limits. Add pauses
   between large batch requests.

3. **Column name mismatches:** Use `TaxaTools::rename_cols()` to standardise
   column names from different sources before stacking.
