---
title: "From Sequences to Standardized Match Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From Sequences to Standardized Match Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, eval = FALSE, purl = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  # purl = FALSE is load-bearing, not redundant with eval = FALSE.
  # R CMD check's "running R code from vignettes" step TANGLES the vignette
  # (knitr::purl()) and sources the result -- purl() honours purl=, not eval=,
  # so without this a documentation-only vignette still executes during check
  # and errors on objects its prose never defines.
  purl = FALSE
)
```

<!--
  EVERY chunk in this vignette carries `eval = FALSE, purl = FALSE` in its own
  header, and that repetition is load-bearing (2026-09-10) -- do not DRY it up
  into the setup chunk. R CMD check's "running R code from vignettes" step
  TANGLES the vignette with knitr::purl(), which does NOT execute the setup
  chunk, so an opts_chunk$set(purl = FALSE) there has no effect at tangle time
  and every chunk is sourced. Same fix as TaxaExpect/vignettes/building-priors.Rmd
  (2026-09-03).
-->

## Purpose

TaxaMatch handles sequence input, BLAST search, and match standardisation.
It produces the canonical "match object" consumed by TaxaLikely (for
likelihood modelling) and TaxaAssign (for posterior computation).

## Quick Start

```{r quickstart, eval = FALSE, purl = FALSE}
library(TaxaMatch)

# From DADA2 output
seqs <- read_sequence_table("seqtab_nochim.rds")
filtered <- filter_sequences(seqs, barcode_term = "12S")
blasted <- blast_sequences(filtered, db = "nt", barcode_term = "12S")
match_df <- standardize_match_data(blasted) |>
  filter_redundant_hypotheses()
```

## Detailed Walkthrough

### Step 1: Read sequence data

```{r read, eval = FALSE, purl = FALSE}
# Option A: DADA2 sequence table (matrix: samples x ASVs)
seqs <- read_sequence_table("seqtab_nochim.rds")

# Option B: FASTA file
seqs <- read_sequence_table("sequences.fasta")

# Option C: With pre-computed taxonomy
seqs <- read_sequence_table("seqtab_nochim.rds",
  taxonomy = "tax_table.rds"
)
```

### Step 2: Filter sequences

```{r filter, eval = FALSE, purl = FALSE}
filtered <- filter_sequences(
  seqs,
  barcode_term  = "12S", # auto-detects length range
  min_abundance = 2 # drop singletons
)
```

### Step 3: BLAST search

```{r blast, eval = FALSE, purl = FALSE}
# Remote NCBI BLAST
blasted <- blast_sequences(
  filtered,
  db           = "nt",
  barcode_term = "12S",
  score_range  = 5 # keep hits within 5% of top score
)

# Local BLAST (requires rBLAST + local database)
# blasted <- blast_sequences(filtered, db = "local_12S.fasta", remote = FALSE)
```

### Step 4: Standardize

```{r standardize, eval = FALSE, purl = FALSE}
match_df <- standardize_match_data(
  blasted,
  observation_id_col = "ESVId",
  score_col = "PercMatch"
)

# Remove redundant higher-rank hypotheses
match_df <- filter_redundant_hypotheses(match_df)
```

## Key Parameters

| Parameter | Default | When to change |
|---|---|---|
| `barcode_term` | — | Must match your marker for length filtering |
| `min_abundance` | 1 | Increase to drop rare/spurious ASVs |
| `score_range` | 5 | Narrow for stricter hit selection |
| `rank_system` | kpcofgs | Subset if your data has fewer ranks |

## Common Pitfalls

1. **BLAST timeout:** Remote NCBI BLAST can be slow for many sequences.
   Use `blast_sequences(..., timeout = 600)` or switch to local BLAST.

2. **DADA2 on Google Drive:** `multithread = TRUE` fails on cloud-synced
   drives. Copy files locally first.

3. **Score column name varies:** Different BLAST tools use different column
   names. `standardize_match_data()` auto-detects common names, or specify
   `score_col` explicitly.
