---
title: "Setting Up API Keys for the TaxaID Ecosystem"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setting Up API Keys for the TaxaID Ecosystem}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

The TaxaID ecosystem uses several external APIs for taxonomic verification,
occurrence data retrieval, habitat assignment, and LLM-powered analysis. This
vignette explains which API keys are needed, where to get them, and how to
configure them.

**Not all keys are required.** Most users only need 2--3, depending on which
packages and workflows they use.

## Quick Reference

| Key | Required by | Free? | Where to get it |
|---|---|---|---|
| `ENTREZ_KEY` | TaxaTools, TaxaLikely, TaxaMatch | Yes | <https://www.ncbi.nlm.nih.gov/account/settings/> |
| `ANTHROPIC_API_KEY` | TaxaTools (LLM functions) | Paid | <https://console.anthropic.com/settings/keys> |
| `GBIF_USER` / `GBIF_PWD` / `GBIF_EMAIL` | TaxaFetch | Yes | <https://www.gbif.org/user/profile> |
| `OPENALEX_API_KEY` | TaxaFetch (literature search) | Yes | <https://openalex.org/settings/api> |
| `GEMINI_API_KEY` | TaxaTools (optional LLM) | Free tier | <https://aistudio.google.com/apikey> |
| `OPENAI_API_KEY` | TaxaTools (optional LLM) | Paid | <https://platform.openai.com/api-keys> |

## Setting Keys in ~/.Renviron

R reads environment variables from `~/.Renviron` at startup. Open it with:

```{r}
usethis::edit_r_environ()
```

Add one key per line (no quotes, no spaces around `=`):

```
ENTREZ_KEY=your_ncbi_api_key_here
ANTHROPIC_API_KEY=sk-ant-your-key-here
GBIF_USER=your_gbif_username
GBIF_PWD=your_gbif_password
GBIF_EMAIL=you@example.com
OPENALEX_API_KEY=your_openalex_key_here
GEMINI_API_KEY=your_gemini_key_here
OPENAI_API_KEY=sk-your-openai-key-here
```

**Restart R after editing** (`Session > Restart R` in RStudio, or quit and
relaunch). Environment variables are only read at startup.

## Verifying Your Keys

After restarting R, verify each key is visible:

```{r}
# Check that keys are set (should print the key value, not "")
Sys.getenv("ENTREZ_KEY")
Sys.getenv("ANTHROPIC_API_KEY")
Sys.getenv("GBIF_USER")
```

### Test an LLM call

```{r}
library(TaxaTools)
call_anthropic_api("Reply with just the word 'connected'.")
# Should return "connected" (or similar)
```

### Test NCBI access

```{r}
# With ENTREZ_KEY set, NCBI allows 10 requests/second (vs 3 without)
rentrez::entrez_search(db = "taxonomy", term = "Fundulus parvipinnis")
```

### Test GBIF access

```{r}
# GBIF credentials are used by rgbif for download requests
rgbif::occ_search(scientificName = "Fundulus parvipinnis", limit = 5)
```

## Key Details

### NCBI Entrez API Key (ENTREZ_KEY)

**Used by:** `verify_taxon_names()` (backbone_id = 4), `fetch_reference_sequences()`,
`audit_barcode_coverage()`, `blast_sequences()`

**Why:** NCBI rate-limits unauthenticated requests to 3/second. With a key, the
limit increases to 10/second. For large taxonomic queries, this makes a significant
difference.

**Get it:**

1. Create an NCBI account at <https://www.ncbi.nlm.nih.gov/account/>
2. Go to Settings (<https://www.ncbi.nlm.nih.gov/account/settings/>)
3. Scroll to "API Key Management" and click "Create an API Key"
4. Copy the key into `~/.Renviron` as `ENTREZ_KEY=...`

The `rentrez` package reads `ENTREZ_KEY` automatically.

### Anthropic API Key (ANTHROPIC_API_KEY)

**Used by:** `call_anthropic_api()` and any function that accepts `llm_fn`
(habitat assignment, LLM-based taxonomic assignment, text generation, expert
review)

**Why:** Claude is the default LLM provider across the TaxaID ecosystem. It
powers habitat assignment (TaxaHabitat), taxonomic priors (TaxaAssign), expert
review (TaxaFlag), and text generation (TaxaTools).

**Get it:**

1. Create an account at <https://console.anthropic.com/>
2. Go to API Keys (<https://console.anthropic.com/settings/keys>)
3. Create a new key and copy it into `~/.Renviron` as `ANTHROPIC_API_KEY=...`

**Cost:** Pay-per-use. Typical TaxaID workflows (habitat assignment for ~100
taxa, taxonomic review) cost under $1 per run with Claude Sonnet.

### GBIF Credentials (GBIF_USER, GBIF_PWD, GBIF_EMAIL)

**Used by:** `fetch_gbif_occurrences()` in TaxaFetch, `build_priors()` in
TaxaExpect

**Why:** GBIF requires authentication for occurrence download requests. Searches
work without credentials, but bulk downloads (which TaxaFetch uses) need a
registered account.

**Get it:**

1. Register at <https://www.gbif.org/user/profile>
2. Add all three variables to `~/.Renviron`

### OpenAlex API Key (OPENALEX_API_KEY)

**Used by:** `search_literature()` in TaxaFetch

**Why:** OpenAlex provides open scholarly metadata. The API works without a key,
but authenticated requests have higher rate limits and are polled. Required since
February 2026.

**Get it:**

1. Visit <https://openalex.org/settings/api>
2. Register for a free key
3. Add to `~/.Renviron` as `OPENALEX_API_KEY=...`

### Alternative LLM Providers (Optional)

TaxaID's `llm_fn` pattern lets you swap LLM providers. These are optional --
only set them if you want to use an alternative to Anthropic.

**Google Gemini (`GEMINI_API_KEY`):**

- Free tier available at <https://aistudio.google.com/apikey>
- Use: `call_gemini_api(prompt, model = "gemini-2.0-flash")`

**OpenAI (`OPENAI_API_KEY`):**

- Paid account at <https://platform.openai.com/api-keys>
- Use: `call_openai_api(prompt, model = "gpt-4o-mini")`

**Ollama (no key needed):**

- Local LLM server: <https://ollama.com/>
- Install Ollama, pull a model (`ollama pull llama3.2`), then:
- Use: `call_ollama_api(prompt, model = "llama3.2")`
- No API key or internet connection needed

### Passing Alternative Providers to TaxaID Functions

Any TaxaID function with an `llm_fn` parameter accepts a provider swap:

```{r}
# Use Gemini instead of Anthropic for habitat assignment
library(TaxaHabitat)
result <- assign_habitat_biological(
  taxa_df,
  llm_fn = TaxaTools::call_gemini_api
)

# Use a local Ollama model (free, private, offline)
result <- assign_habitat_biological(
  taxa_df,
  llm_fn = function(prompt) TaxaTools::call_ollama_api(prompt, model = "llama3.2")
)
```

## Troubleshooting

**"Error: API key not found"** -- The key is not in your environment. Check
`Sys.getenv("KEY_NAME")`. If empty, edit `~/.Renviron` and restart R.

**"Error: 401 Unauthorized"** -- The key is set but invalid. Regenerate it at
the provider's website.

**"Error: 429 Too Many Requests"** -- Rate limited. Wait a moment and retry.
For NCBI, ensure `ENTREZ_KEY` is set (increases limit from 3 to 10 req/sec).

**LLM responses are truncated** -- Some providers have low default `max_tokens`.
Pass a higher value: `call_anthropic_api(prompt, max_tokens = 4096)`.

**Ollama connection refused** -- Ensure the Ollama server is running
(`ollama serve` in a terminal).
