---
title: "Cleaning and Verifying Taxon Names"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cleaning and Verifying Taxon Names}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Purpose

TaxaTools provides name verification, cleaning, parsing, and standardisation
utilities used throughout the TaxaID ecosystem. It also hosts the LLM provider
functions and text generation tools.

## Quick Start

```{r quickstart}
library(TaxaTools)

# Clean raw names
clean <- clean_taxon_names(c("Fundulus parvipinnis (Girard, 1854)",
                              "Atherinops  affinis", "sp. nov."))
# "Fundulus parvipinnis", "Atherinops affinis", NA

# Verify against a backbone
verified <- verify_taxon_names(c("Fundulus parvipinnis", "Atherinops afinis"),
                                backbone_id = 4)
```

## Core Functions

### Name cleaning

`clean_taxon_names()` removes authors, subspecies, formatting artifacts, and
invalid entries. Output is the same length as input; invalid names become `NA`.

```{r clean}
raw <- c("Fundulus parvipinnis (Girard, 1854)",
         "Atherinops  affinis",
         "Uncultured bacterium clone",
         "Cottus sp.",
         "")
clean_taxon_names(raw)
# "Fundulus parvipinnis", "Atherinops affinis", NA, NA, NA
```

### Name verification

`verify_taxon_names()` checks names against the Global Names Verifier API.

```{r verify}
verified <- verify_taxon_names(
  c("Fundulus parvipinnis", "Atherinops afinis"),  # note typo
  backbone_id = 4  # NCBI
)
# Returns: matched_name, classification_path, match_type, etc.
```

### Backbone translation

`change_backbone()` translates names between taxonomic backbones (GBIF, NCBI,
WoRMS, etc.).

```{r backbone}
translated <- verified |>
  change_backbone(input_col = "user_supplied_name",
                  old_backbone_label = "gbif_name",
                  new_backbone_label = "ncbi_name")
```

### Creating taxon names from taxonomy columns

```{r create}
df <- data.frame(family = "Fundulidae", genus = "Fundulus",
                 species = "Fundulus parvipinnis")
create_taxon_names(df, rank_system = c("family", "genus", "species"))
# Adds: taxon_name = "Fundulus parvipinnis", taxon_name_rank = "species"
```

### Utility functions

```{r utils}
# Detect rank columns in a data frame
detect_ranks(match_df)

# Validate species binomials
is_plausible_binomial(c("Cottus asper", "Cottus sp.", "uncultured clone"))
# TRUE, FALSE, FALSE

# Resolve barcode length ranges
resolve_barcode_lengths("MiFishU")  # c(100, 600)

# Find taxonomy conflicts
find_taxonomy_conflicts(taxonomy_df)
```

## LLM Provider Functions

TaxaTools hosts the `llm_fn` pattern used across the ecosystem:

```{r llm}
# call_api() auto-selects the provider from your API keys (set at library load)
result <- call_api("What habitat does Fundulus parvipinnis prefer?")

# Inspect token usage
attr(result, "tokens")  # list(input = N, output = N)

# Print token counts inline (useful during prompt development)
result <- call_api("What habitat does Fundulus parvipinnis prefer?",
                   show_tokens = TRUE)

# Guard against large prompts before making the HTTP call
result <- call_api(big_prompt, max_input_tokens = 5000)

# Explicit provider or tier
result <- call_api(prompt, provider = "gemini", tier = "fast")
result <- call_api(prompt, provider = "openai", model = "gpt-4o-mini")

# Provider wrapper functions are still available for explicit control
result <- call_anthropic_api(prompt)
result <- call_ollama_api(prompt, model = "llama3.2")
```

## Common Pitfalls

1. **`clean_taxon_names()` returns NA:** This is intentional for invalid
   names. Use `na.omit()` to filter: `clean_taxon_names(x) |> na.omit()`.

2. **Backbone ID confusion:** NCBI = 4, GBIF = 11, WoRMS = 9. See
   `?verify_taxon_names` for the full list.

3. **API key not set:** LLM functions require API keys in `~/.Renviron`
   (e.g., `ANTHROPIC_API_KEY`, `GEMINI_API_KEY`).
